### 2024-09-17 paper/paper.qmd
initial paper commit

---
title: "My title"
subtitle: "My subtitle if needed"
author: 
  - Lexun Yu
thanks: "Code and data are available at: [https://github.com/yulexun/toronto-fire](https://github.com/yulexun/toronto-fire)."
date: today
date-format: long
abstract: "First sentence. Second sentence. Third sentence. Fourth sentence."
format: pdf
number-sections: true
bibliography: bibliography.bib
---

# Introduction
This project uses @citeR.

# Bibliography

### 2024-09-17 paper/paper.qmd
import data to paper

@@
 format: pdf
+toc: true
 number-sections: true
@@
 
-# Introduction
+```{r}
+#| include: false
+#| warning: false
+#| message: false
+
+#### Workspace setup ####
+library(tidyverse)
+library(tinytex)
+library(dplyr)
+library(tibble)
+library(opendatatoronto)
+library(janitor)
+library(ggplot2)
+library(knitr)
+library(here)
+
+# # Set the number of decimal places to 2 for all subsequent figures and tables.
+# options(digits = 2)
+
+# Read data, https://stackoverflow.com/questions/78894145/quarto-document-not-seeing-project-directory-for-targets-data-functions-or-scr
+raw_lead_data <- read_csv(here("data/raw/toronto_fire_incidents.csv"))
+cleaned_data <- read_csv(here("data/analysis/cleaned_fire_data_simple.csv"))
+```
+
+# Introduction {#sec-intro}
 This project uses @citeR.
 
+\newpage
 # Bibliography

### 2024-09-18 paper/paper.qmd
partially finished introduction

@@
 # Introduction {#sec-intro}
+The urban fire hazard is one of the most pressing issues in this context, especially in Canada where cities are dealing with such issues like climate, facilities or population density.As of July 1 2023, the population in urban area in Canada reached 33,812,133 [@statsCanada].
+Not only do fires in highly populated regions result in heavy losses in terms of property, but also, in the contest of people and the environment, the consequences are enormous.
+Also, urban fires are resources-dependent and require attention from city services and emergency services, thus indicating the need for prevention, action and planning based on risk assessment.
+With such issues in mind, the knowledge of urban fire hazards in Canadian cities is needed for creating policies that address public safety and increase urban resilience.
+
+In Canada, articles about fire incidents has a focus on wildfire.
+For instance, @Goemans2012 discuss the fire mitigation plan during the 2003 wildfire at Kelowna, British Columbia, while @Mamuji2018 talk about the evacuation during the Fort McMurray wildfire in Alberta.
+The researches conducted about urban fire incidents are done in other parts of the world such as East Asia.
+@MasoodRafi2012 research the nature and level of this threat.
+They conclude that the lack of training in fire department, shortage of facilities and infrastructure the major issues in Pakistan.
+The research by @HariMurti2023 in Semarang City also emphasize the importance of community understanding and the installation of fire protection facilities.
+This article uses the data provided by opendatatoronto library [@opendatatoronto] in order to analyze fire occurrences in the city of Toronto, which is an important research gap in the study of fire incidents in America.
+This study seeks to provide a deeper understanding of fire patterns to improve fire prevention and emergency response strategies.
+
+In this paper we visualize Toronto's Fire Incidents' data.
+
 This project uses @citeR.

### 2024-09-18 paper/paper.qmd
change read data

@@
 # Read data, https://stackoverflow.com/questions/78894145/quarto-document-not-seeing-project-directory-for-targets-data-functions-or-scr
-raw_lead_data <- read_csv(here("data/raw/toronto_fire_incidents.csv"))
+raw_data <- read_csv(here("data/raw/toronto_fire_incidents.csv"))
 cleaned_data <- read_csv(here("data/analysis/cleaned_fire_data_simple.csv"))

### 2024-09-18 paper/paper.qmd
add sketches
add quote
add two tables in cleaned data

@@
 # # Set the number of decimal places to 2 for all subsequent figures and tables. 
+# # https://github.com/AbbassSleiman/Lead-Concentrations/blob/main/outputs/paper/lead_concentration.qmd
 # options(digits = 2)
@@
 cleaned_data <- read_csv(here("data/analysis/cleaned_fire_data_simple.csv"))
+cleaned_data_tfs <-
+  cleaned_data |>
+  select(
+    tfs_alarm_time, 
+    tfs_arrival_time, 
+    time_diff,
+    civilian_casualties, 
+    estimated_dollar_loss,
+  )
+cleaned_data_cat <-
+  cleaned_data |>
+  select(
+    area_of_origin, 
+    ignition_source, 
+    fire_alarm_system_presence, 
+    smoke_alarm_at_fire_origin, 
+    sprinkler_system_operation
+  )
 ```
@@
 
-This project uses @citeR.
+# Data
+
+## Raw Data
+This project uses @citeR, @opendatatoronto, @knitr, @tidyverse, @ggplot2
+
+## Cleaned Data
+
+```{r}
+#| label: table-clean-tfs
+#| tbl-cap: Top rows of cleaned Toronto Fire Service response time and loss data
+#| echo: false
+
+head(cleaned_data_tfs, 5) |>
+  kable(
+    col.names = c("Alarm Time", "TFS Arrival Time", "TFS Response Time"
+    , "Civilian Casualties", "Estimated Loss in Dollar"),
+      )
+```
+
+```{r}
+#| label: table-clean-cat
+#| tbl-cap: Top rows of cleaned data showing Area of Origin, Ignition source and Fire, Smoke, Sprnkler System Presence.
+#| echo: false
+
+head(cleaned_data_cat, 5) |>
+  kable(
+    col.names = c("Area of Origin", "Ignition Source", "Fire Alarm Status"
+    , "Smoke Alarm Status", "Sprinkler System Status"),
+      )
+cat("Note: 'PO' = System present, \n'N' = System not present, \n'P' = System present but not operated")
+```
 

### 2024-09-18 paper/paper.qmd
add attribution

@@
   )
+
+legend <- read.csv(here("data/legend.txt"), header = TRUE, stringsAsFactors = FALSE)
 ```
@@
 ## Raw Data
-This project uses @citeR, @opendatatoronto, @knitr, @tidyverse, @ggplot2
+This project uses @citeR, @opendatatoronto, @knitr, @tidyverse, @ggplot2, @dplyr
 
 ## Cleaned Data
+
+```{r}
+#| echo: false
+#| 
+# Convert str to factor
+
+cleaned_data_cat <-
+  cleaned_data_cat %>%
+  mutate_at(vars(
+    area_of_origin,
+    fire_alarm_system_presence, ignition_source,
+    smoke_alarm_at_fire_origin, sprinkler_system_operation
+  ), as.factor)
+```
 
@@
 \newpage
+
+\appendix
+
+# Appendix {#sec-appendix}
+
+## Graph of areas of origin of fire incidents by number of occurrences {#sec-origin-fig}
+```{r, fig.width=10,fig.height=11}
+#| label: fig-origin
+#| fig-cap: Areas of origin of fire incidents by number of occurrences
+#| echo: false
+
+cleaned_data_cat |>
+  count(area_of_origin) |>
+  ggplot(mapping = aes(x = n, y = area_of_origin)) +
+  geom_col() + 
+  theme_minimal() + 
+  labs(x = "number of observation", y = "area of origin")
+
+cat("Index of 'Area of Origin':")
+cat(legend$x, sep = "\n")
+```
+
+# Attribution Statement
+Contains information licensed under the Open Government Licence – Toronto. 
+Visit: [https://open.toronto.ca/open-data-license/](https://open.toronto.ca/open-data-license/)
+
+\newpage
 # Bibliography

### 2024-09-18 scripts/02-data_cleaning.r

@@ -161,6 +161,28 @@ cleaned_fire_data_simple <-
       )
   )
 
+# keep only the indicator number of area_of_origin,
+# extent of fire and ignition source
+cleaned_fire_data_simple <-
+  cleaned_fire_data_simple |>
+  mutate(
+    area_of_origin =
+      substr(area_of_origin, start = 1, stop = 2)
+  )
+
+cleaned_fire_data_simple <-
+  cleaned_fire_data_simple |>
+  mutate(
+    extent_of_fire =
+      substr(extent_of_fire, start = 1, stop = 2)
+  )
+
+cleaned_fire_data_simple <-
+  cleaned_fire_data_simple |>
+  mutate(
+    ignition_source =
+      substr(ignition_source, start = 1, stop = 2)
+  )
 
 # Write the cleaned data to a new csv
 write_csv(

### 2024-09-18 scripts/05-legend_for_figures.r

@@ -1,0 +1,18 @@
+#### Preamble ####
+# Purpose: Generate legend for categorial figures
+# Author: Lexun Yu
+# Date: 16 Sep 2024
+# Contact: lx.yu@mail.utoronto.ca
+# License: MIT
+# Pre-requisites: None
+
+cleaned_data <- 
+  read_csv("data/analysis/cleaned_fire_data.csv", show_col_types = FALSE)
+
+data <- 
+  cleaned_data$area_of_origin |>
+  unique() |>
+  sort.default()
+
+write.table(data, file = "data/legend.txt", sep = "\n",
+            row.names = FALSE)

### 2024-09-19 paper/paper.qmd
add paper data section

--- 2024-09-18 paper/paper.qmd
@@
 cleaned_data <- read_csv(here("data/analysis/cleaned_fire_data_simple.csv"))
+cleaned_data_tfs <-
+  cleaned_data |>
+  select(
+    tfs_alarm_time, 
+    tfs_arrival_time, 
+    time_diff,
+    civilian_casualties, 
+    estimated_dollar_loss,
+  )
+
+cleaned_data_cat <-
+  cleaned_data |>
+  select(
+    estimated_dollar_loss,
+    area_of_origin, 
+    ignition_source, 
+    fire_alarm_system_presence, 
+    smoke_alarm_at_fire_origin, 
+    sprinkler_system_operation
+  )
+
+legend_origin <- read.csv(here("data/legend/area_of_origin.txt"), header = TRUE, stringsAsFactors = FALSE)
+legend_source <- read.csv(here("data/legend/ignition_source.txt"), header = TRUE, stringsAsFactors = FALSE)
+```
+
+```{r}
+#| echo: false
+# Convert str to factor
+
+cleaned_data_cat <-
+  cleaned_data_cat %>%
+  mutate_at(vars(
+    area_of_origin,
+    fire_alarm_system_presence, ignition_source,
+    smoke_alarm_at_fire_origin, sprinkler_system_operation
+  ), as.factor)
 ```
@@
 
-In this paper we visualize Toronto's Fire Incidents' data.
+In this paper we visualize Toronto's Fire Incidents data.
 
-This project uses @citeR.
+# Data
+
+## Overview {#sec-data-overview}
+
+<!-- Referenced https://github.com/AbbassSleiman/Lead-Concentrations/blob/main/outputs/paper/lead_concentration.pdf for formatting. -->
+The data used in this paper is obtained from the opendatatoronto library [@opendatatoronto].
+The dataset used is Fire Incidents.
+According to @opendatatoronto, it includes fire incidents as defined by the Ontario Fire Marshal (OFM) up to December 31, 2023.
+The data gathering and analysis is done in R [@citeR] with the following packages: opendatatoronto [@opendatatoronto], knitr [@knitr], tidyverse [@tidyverse], ggplot2 [@ggplot2], dplyr [@dplyr], and lubridate [@lubridate].
+
+The cleaned data are divided into two groups in accordance to the two most important factors in urban fire incidents: responsiveness of fire department and fire protection equipments, as indicated by @HariMurti2023 and @MasoodRafi2012.
+The first group focuses on the fire services' response time and the loss of life or money, and the second group focuses on the loss of money, reason of fire incidents and the presence or operation of fire protection equipments.
+All other data features are ignored during the data cleaning process.
+
+The first group of data shows the responsiveness of Toronto Fire Services.
+An example of this dataset is presented in @tbl-clean-tfs.
+"Alarm Time" is the time when TFS are notified of the incident.
+"TFS Arrival time" is the timestamp of first arriving unit.
+The difference in minute is calculated at the data cleaning step as "TFS Response Time".
+The number under "Civilian Casualties" indicates the fire related injury or fatality.
+"Estimated Loss in Dollar" is the estimated loss measured in dollar.
+
+```{r}
+#| label: tbl-clean-tfs
+#| tbl-cap: Top rows of cleaned Toronto Fire Service response time and loss data
+#| echo: false
+
+head(cleaned_data_tfs, 5) |>
+  kable(
+    col.names = c("Alarm Time", "TFS Arrival Time", "TFS Response Time"
+    , "Civilian Casualties", "Estimated Loss in Dollar"),
+      )
+```
+
+The second group of data shows the money loss, cause of fire and the presence of fire prevention equipments.
+@tbl-clean-cat shows the first five rows of the second group of data.
+"Estimated Loss in Dollar" is the estimated loss measured in dollar.
+"Area of origin" indicates the area fire originate.
+"Ignition Source" shows the object causing fire.
+"Fire Alarm Status", "Smoke Alarm Status" and "Sprinkler System Status" indicates the presence and operation of fire protection equipments.
+
+```{r}
+#| label: tbl-clean-cat
+#| tbl-cap: Top rows of cleaned data showing Area of Origin, Ignition source and Fire, Smoke, Sprnkler System Presence.
+#| echo: false
+
+head(cleaned_data_cat, 5) |>
+  kable(
+    col.names = c("Estimated Loss in Dollar", "Area of Origin", "Ignition Source", "Fire Alarm Status"
+    , "Smoke Alarm Status", "Sprinkler System Status"),
+      )
+
+cat("Note: 'PO' = System present, \n'N' = System not present, \n'P' = System present but not operated")
+```
+
+## Toronto Fire Service Responsiveness
+
+```{r}
+#| label: fig-clean-tfs
+#| fig-cap: TFS response time and estimated dollar loss over time
+#| echo: false
+#| fig-subcap: ["TFS response time over time","Estimated dollar loss over time"]
+#| layout-ncol: 2
+
+cleaned_data_tfs |>
+  ggplot(mapping = aes(x = tfs_alarm_time, y = time_diff)) + 
+  geom_line() +
+  theme_minimal() +
+  labs(
+    x = "Alarm Time", y = "Response Time",
+  )
+
+cleaned_data_tfs |>
+  ggplot(mapping = aes(x = tfs_alarm_time, y = estimated_dollar_loss)) + 
+  geom_line() +
+  theme_minimal() +
+  labs(
+    x = "Alarm Time", y = "Estimated dollar loss",
+  )
+```
+
+## Fire Protection Equipments
+
+\newpage
+
+\appendix
+
+# Appendix {#sec-appendix}
+
+## Graph of areas of origin of fire incidents by number of occurrences {#sec-origin-fig}
+```{r, fig.width=10,fig.height=11}
+#| label: fig-origin
+#| fig-cap: Areas of origin of fire incidents by number of occurrences
+#| echo: false
+
+cleaned_data_cat |>
+  count(area_of_origin) |>
+  ggplot(mapping = aes(x = n, y = area_of_origin)) +
+  geom_col() + 
+  theme_minimal() + 
+  labs(x = "number of observation", y = "area of origin")
+
+cat("Index of 'Area of Origin':")
+cat(legend_origin$x, sep = "\n")
+```
+
+\newpage
+
+## Graph of ignition source of fire incidents by number of occurrences {#sec-source-fig}
+
+```{r, fig.width=10,fig.height=11}
+#| label: fig-source
+#| fig-cap: Ignition Source of fire incidents by number of occurrences
+#| echo: false
+
+cleaned_data_cat |>
+  count(ignition_source) |>
+  ggplot(mapping = aes(x = n, y = ignition_source)) +
+  geom_col() + 
+  theme_minimal() + 
+  labs(x = "number of observation", y = "ignition source")
+
+cat("Index of 'Ignition Source':")
+cat(legend_source$x, sep = "\n")
+```
+
+## Attribution Statement
+Contains information licensed under the Open Government Licence – Toronto. 
+Visit: [https://open.toronto.ca/open-data-license/](https://open.toronto.ca/open-data-license/)
 

### 2024-09-19 scripts/05-legend_for_figures.r

@@ -14,5 +14,13 @@ data <-
   unique() |>
   sort.default()
 
-write.table(data, file = "data/legend.txt", sep = "\n",
+write.table(data, file = "data/legend/area_of_origin.txt", sep = "\n",
+            row.names = FALSE)
+
+data <- 
+  cleaned_data$ignition_source |>
+  unique() |>
+  sort.default()
+
+write.table(data, file = "data/legend/ignition_source.txt", sep = "\n",
             row.names = FALSE)

### 2024-09-20 paper/paper.qmd
change paper

@@
 
-The first group of data shows the responsiveness of Toronto Fire Services.
-An example of this dataset is presented in @tbl-clean-tfs.
-"Alarm Time" is the time when TFS are notified of the incident.
-"TFS Arrival time" is the timestamp of first arriving unit.
-The difference in minute is calculated at the data cleaning step as "TFS Response Time".
-The number under "Civilian Casualties" indicates the fire related injury or fatality.
-"Estimated Loss in Dollar" is the estimated loss measured in dollar.
-
-```{r}
-#| label: tbl-clean-tfs
-#| tbl-cap: Top rows of cleaned Toronto Fire Service response time and loss data
-#| echo: false
-
-head(cleaned_data_tfs, 5) |>
-  kable(
-    col.names = c("Alarm Time", "TFS Arrival Time", "TFS Response Time"
-    , "Civilian Casualties", "Estimated Loss in Dollar"),
-      )
-```
-
 The second group of data shows the money loss, cause of fire and the presence of fire prevention equipments.
@@
 
+The first group of data shows the responsiveness of Toronto Fire Services.
+An example of this dataset is presented in @tbl-clean-tfs.
+"Alarm Time" is the time when TFS are notified of the incident.
+"TFS Arrival time" is the timestamp of first arriving unit.
+The difference in minute is calculated at the data cleaning step as "TFS Response Time".
+The number under "Civilian Casualties" indicates the fire related injury or fatality.
+"Estimated Loss in Dollar" is the estimated loss measured in dollar.
+
 ```{r}
-#| label: fig-clean-tfs
+#| label: tbl-clean-tfs
+#| tbl-cap: Top rows of cleaned Toronto Fire Service response time and loss data
+#| echo: false
+
+head(cleaned_data_tfs, 5) |>
+  kable(
+    col.names = c("Alarm Time", "TFS Arrival Time", "TFS Response Time"
+    , "Civilian Casualties", "Estimated Loss in Dollar"),
+      )
+```
+
+@fig-clean-tfs-diff-dollar-time-1 shows the line plot of Toronto Fire Service's response time over the period between 2013 to 2023.
+
+```{r}
+#| label: fig-clean-tfs-diff-dollar-time
 #| fig-cap: TFS response time and estimated dollar loss over time
@@
   )
+
+```
+
+```{r}
+#| label: fig-clean-tfs-boxplot
+#| fig-cap: "Boxplot of TFS response time data and estimated dollar loss"
+#| fig-subcap: ["Boxplot of TFS response time", "Boxplot of estimated dollar loss"]
+#| echo: false
+#| layout-ncol: 2
+
+cleaned_data_tfs |>
+  ggplot(aes(x = "", y = time_diff)) +
+  geom_boxplot() +
+  labs(y = "TFS response time", x = "") + 
+  theme_classic()
+
+cleaned_data_tfs |>
+  ggplot(aes(x = "", y = estimated_dollar_loss)) +
+  geom_boxplot() +
+  labs(y = "Estimated dollar loss", x = "") + 
+  theme_classic()
+
+```
+
+```{r}
+#| label: tbl-clean-tfs-summary
+#| tbl-cap: "Summary Statistic of TFS response time data and estimated dollar loss"
+#| tbl-subcap: ["Summary of TFS response time", "Summary of estimated dollar loss"]
+#| echo: false
+#| layout-ncol: 2
+
+summary_response_time <- 
+  cleaned_data_tfs$time_diff |>
+  summary() |> append(sd(cleaned_data_tfs$time_diff))
+
+summary_response_time <- data.frame(
+  Statistic = c(names(summary(cleaned_data_tfs$time_diff)), "SD"),
+  Value = as.vector(summary_response_time)
+)
+
+kable(summary_response_time, col.names = c("Statistic", "Value"))
+
+summary_dollar_loss <- 
+  cleaned_data_tfs$estimated_dollar_loss |>
+  summary() |> append(sd(cleaned_data_tfs$estimated_dollar_loss))
+
+summary_dollar_loss <- data.frame(
+  Statistic = c(names(summary(cleaned_data_tfs$estimated_dollar_loss)), "SD"),
+  Value = as.vector(summary_dollar_loss)
+)
+
+kable(summary_dollar_loss, col.names = c("Statistic", "Value"))
+
 ```

### 2024-09-20 paper/paper.qmd
delete civilian casualties

@@
     time_diff,
-    civilian_casualties, 
     estimated_dollar_loss,
@@
 
-The second group of data shows the money loss, cause of fire and the presence of fire prevention equipments.
-@tbl-clean-cat shows the first five rows of the second group of data.
-"Estimated Loss in Dollar" is the estimated loss measured in dollar.
-"Area of origin" indicates the area fire originate.
-"Ignition Source" shows the object causing fire.
-"Fire Alarm Status", "Smoke Alarm Status" and "Sprinkler System Status" indicates the presence and operation of fire protection equipments.
-
-```{r}
-#| label: tbl-clean-cat
-#| tbl-cap: Top rows of cleaned data showing Area of Origin, Ignition source and Fire, Smoke, Sprnkler System Presence.
-#| echo: false
-
-head(cleaned_data_cat, 5) |>
-  kable(
-    col.names = c("Estimated Loss in Dollar", "Area of Origin", "Ignition Source", "Fire Alarm Status"
-    , "Smoke Alarm Status", "Sprinkler System Status"),
-      )
-
-cat("Note: 'PO' = System present, \n'N' = System not present, \n'P' = System present but not operated")
-```
-
 ## Toronto Fire Service Responsiveness
@@
 The difference in minute is calculated at the data cleaning step as "TFS Response Time".
-The number under "Civilian Casualties" indicates the fire related injury or fatality.
 "Estimated Loss in Dollar" is the estimated loss measured in dollar.
@@
   kable(
-    col.names = c("Alarm Time", "TFS Arrival Time", "TFS Response Time"
-    , "Civilian Casualties", "Estimated Loss in Dollar"),
+    col.names = c("Alarm Time", "TFS Arrival Time", "TFS Response Time", "Estimated Loss in Dollar"),
       )
@@
 @fig-clean-tfs-diff-dollar-time-1 shows the line plot of Toronto Fire Service's response time over the period between 2013 to 2023.
+The response time fluctuated around a relatively consistent level, typically between 5 and 10 minutes, with occasional spikes where the response time increased significantly above the usual range, particularly around 2014 and 2019.
+@fig-clean-tfs-diff-dollar-time-2 shows the line plot of estimated dollar loss of the fire incidents.
+Most of the incidents resulted in relatively low dollar losses, but there are several spikes that represent incidents where the loss was significantly higher.
+Some of the most notable spikes occurred between 2015 and 2019, with losses exceeding 2 million dollars.
 
@@
   )
+```
 
-```
+@tbl-clean-tfs-summary-1 indicates the summary statistic and @fig-clean-tfs-boxplot-1 graphs the boxplot of TFS's response time.
+The value mainly range from 4.03 minutes to 5.7 minutes as shown in @tbl-clean-tfs-summary-1, with a mean value of 4.911868 minutes.
+Notably, in @fig-clean-tfs-boxplot-1, several outliers are indicated by dots above the upper whisker, suggesting occasional instances of significantly higher response times.
+
+@tbl-clean-tfs-summary-2 indicates the summary statistic and @fig-clean-tfs-boxplot-2 depicts the distribution of financial losses in boxplot.
+In @tbl-clean-tfs-summary-2, values span from 0 dollar to over 3000000 dollars, with median value of 3000 dollars, mean value of 24551.64 dollars and standard deviation of 93520.08 dollars, indicating a skewed distribution.
+@fig-clean-tfs-boxplot-2 reveals numerous outliers above the upper whisker, highlighting instances of exceptionally high financial losses.
 
@@
 
+The second group of data shows the money loss, cause of fire and the presence of fire prevention equipments.
+@tbl-clean-cat shows the first five rows of the second group of data.
+"Estimated Loss in Dollar" is the estimated loss measured in dollar.
+"Area of origin" indicates the area fire originate.
+"Ignition Source" shows the object causing fire.
+"Fire Alarm Status", "Smoke Alarm Status" and "Sprinkler System Status" indicates the presence and operation of fire protection equipments.
+In the second group of data, 'PO' = System present, 'N' = System not present, and 'P' = System present but not operated.
+
+```{r}
+#| label: tbl-clean-cat
+#| tbl-cap: Top rows of cleaned data showing Area of Origin, Ignition source and Fire, Smoke, Sprnkler System Presence.
+#| echo: false
+
+head(cleaned_data_cat, 5) |>
+  kable(
+    col.names = c("Estimated Loss in Dollar", "Area of Origin", "Ignition Source", "Fire Alarm Status"
+    , "Smoke Alarm Status", "Sprinkler System Status"),
+      )
+```
+
 \newpage

### 2024-09-20 scripts/00-simulate.r

@@ -50,7 +50,6 @@ smoke_alarm_at_fire_origin <- sample(c("PO", "N", "p"), n, replace = TRUE)
 sprinkler_system_operation <- sample(c("P", "N", "PO"), n, replace = TRUE)
 
 # Simulate numerical columns
-civilian_casualties <- sample(0:5, n, replace = TRUE)
 estimated_dollar_loss <- sample(0:60000, n, replace = TRUE)
 
 # Create the simulated dataframe
@@ -58,7 +57,6 @@ simulated_data <- data.frame(
   tfs_alarm_time = alarm_times,
   tfs_arrival_time = arrival_times,
   area_of_origin = area_of_origin,
-  civilian_casualties = civilian_casualties,
   estimated_dollar_loss = estimated_dollar_loss,
   extent_of_fire = extent_of_fire,
   fire_alarm_system_presence = fire_alarm_system_presence,

### 2024-09-20 scripts/02-data_cleaning.r

@@ -29,7 +29,6 @@ selected_cleaned_fire_data <-
     tfs_alarm_time,
     tfs_arrival_time,
     area_of_origin,
-    civilian_casualties,
     estimated_dollar_loss,
     extent_of_fire,
     fire_alarm_system_presence,
@@ -73,11 +72,10 @@ selected_cleaned_fire_data <- selected_cleaned_fire_data %>%
     ., "9 - Activation/operation undetermined"
   )))
 
-# Convert columns 'civilian_casualties'
-# and 'estimated_dollar_loss' back to numeric
+# Convert columns
+# 'estimated_dollar_loss' back to numeric
 selected_cleaned_fire_data <- selected_cleaned_fire_data %>%
   mutate(
-    civilian_casualties = as.numeric(civilian_casualties),
     estimated_dollar_loss = as.numeric(estimated_dollar_loss)
   ) %>%
   drop_na() # Remove rows with NA values

### 2024-09-20 scripts/04-test.r

@@ -21,38 +21,29 @@ test1 <-
 
 print(test1)
 
-# test2: Tests that there are 70 unique area of fire hazard origin
+# test3: test if the
+# estimated dollar loss are non-negative
 test2 <-
-  cleaned_data$area_of_origin |>
-    unique() |>
-    length() == 70
+  cleaned_data$estimated_dollar_loss |> min() >= 0
 
 print(test2)
 
-# test3: test if the cimivilian casualties and
-# estimated dollar loss are non-negative
+# test4, 5, 6: test if the fire alarm,
+# smoke and sprinkler system value is simplified
 test3 <-
-  cleaned_data$civilian_casualties |> min() >= 0
-cleaned_data$estimated_dollar_loss |> min() >= 0
+  cleaned_data$fire_alarm_system_presence |>
+    unique() == c("P", "N")
 
 print(test3)
 
-# test4, 5, 6: test if the fire alarm,
-# smoke and sprinkler system value is simplified
 test4 <-
-  cleaned_data$fire_alarm_system_presence |>
-    unique() == c("P", "N")
+  cleaned_data$smoke_alarm_at_fire_origin |>
+    unique() == c("PO", "N", "P")
 
 print(test4)
 
 test5 <-
-  cleaned_data$smoke_alarm_at_fire_origin |>
-    unique() == c("PO", "N", "P")
-
-print(test5)
-
-test6 <-
   cleaned_data$sprinkler_system_operation |>
     unique() == c("P", "N", "PO")
 
-print(test6)
+print(test5)

### 2024-09-20 scripts/05-legend_for_figures.r

@@ -6,21 +6,25 @@
 # License: MIT
 # Pre-requisites: None
 
-cleaned_data <- 
+cleaned_data <-
   read_csv("data/analysis/cleaned_fire_data.csv", show_col_types = FALSE)
 
-data <- 
+data <-
   cleaned_data$area_of_origin |>
   unique() |>
   sort.default()
 
-write.table(data, file = "data/legend/area_of_origin.txt", sep = "\n",
-            row.names = FALSE)
+write.table(data,
+  file = "data/legend/area_of_origin.txt", sep = "\n",
+  row.names = FALSE
+)
 
-data <- 
+data <-
   cleaned_data$ignition_source |>
   unique() |>
   sort.default()
 
-write.table(data, file = "data/legend/ignition_source.txt", sep = "\n",
-            row.names = FALSE)
+write.table(data,
+  file = "data/legend/ignition_source.txt", sep = "\n",
+  row.names = FALSE
+)

### 2024-09-21 paper/paper.qmd
finished data

@@
     estimated_dollar_loss,
+    extent_of_fire
   )
@@
   kable(
-    col.names = c("Alarm Time", "TFS Arrival Time", "TFS Response Time", "Estimated Loss in Dollar"),
+    col.names = c("Alarm Time", "TFS Arrival Time", "TFS Response Time", "Estimated Loss in Dollar", "Extent of Fire"),
       )
@@
 
+@fig-bar-extent illustrates that most of the observed fires are relatively small in size.
+The x-axis depicts the extent of the fire, while the y-axis shows the frequency of observations.
+A notably tall bar at the start signifies a high occurrence of smaller fires.
+As fire size increases along the x-axis, the number of observations drops sharply, indicating that larger fires are far less frequent.
+
+```{r}
+#| label: fig-bar-extent
+#| fig-cap: "Distribution of extent of fire"
+#| echo: false
+
+cleaned_data_tfs |>
+  count(extent_of_fire) |>
+  ggplot(mapping = aes(x = extent_of_fire, y = n)) +
+  geom_col() + 
+  theme_minimal() + 
+  labs(x = "extent of fire", y = "number of observation")
+```
+
+@fig-response-vs-money shows a line of best fit, which indicates a weak correlation between TFS response time and estimated dollar loss.
+There is a wide range of TFS response times for lower dollar losses.
+```{r}
+#| label: fig-response-vs-money
+#| fig-cap: "Line of best fit between TFS response time and dollar loss"
+#| echo: false
+#| warning: false
+#| message: false
+
+cleaned_data_tfs |>
+  ggplot(aes(x = estimated_dollar_loss, y = time_diff)) +
+  geom_jitter() +
+  geom_smooth() +
+  theme_minimal() +
+  labs(x = "Estimated dollar loss", y = "TFS response time")
+
+```
+
 ## Fire Protection Equipments
@@
 ```
+
+@fig-origin in [Appendix A.1](#sec-origin-fig) illustrates the distribution of the fire incidents' area of origin.
+It consists 71 unique categories.
+The most common area of origin is category 24, which is cooking area or kitchen.
+Other common areas include category 64: Porch or Balcony, 22: Sleeping Area or Bedroom, 21: Living Area, and 27: Laundry Area.
+However, fires in these locations occur significantly less often compared to those in kitchens.
+
+@fig-source in [Appendix A.2](#sec-source-fig) shows the distribution of ignition source in fire incidents.
+It contains 73 unique categories.
+The most common ignition source is 11: Stove, Range-top burner and 71: Smoker's Articles.
+Other common ignition sources include 55: Candle, 43: Clothes Dryer, and 12: Oven.
+
+```{r}
+#| echo: false
+
+# create dollar loss group
+
+cleaned_data_cat <-
+  cleaned_data_cat |>
+  mutate(
+    categorized_loss =
+      case_when(
+        estimated_dollar_loss < 500 ~ "<500",
+        estimated_dollar_loss < 3000 ~ "500-2,999",
+        estimated_dollar_loss < 15000 ~ "3,000-14,999",
+        estimated_dollar_loss < 100000 ~ "15,000-99,999",
+        estimated_dollar_loss < 3000001 ~ "100,000-3,000,000"
+      ),
+    categorized_loss = 
+      factor(categorized_loss, levels = c("<500", "500-2,999", "3,000-14,999", "15,000-99,999", "100,000-3,000,000"))
+  )
+```
+
+```{r}
+#| label: fig-financial-loss-group
+#| fig-cap: "Distribution of financial loss group, and fire prevention system presence"
+#| fig-subcap: ["Distribution of financial loss group, and fire alarm presence", "Distribution of financial loss group, and smoke alarm presence", "Distribution of financial loss group, and sprinkler system presence"]
+#| echo: false
+#| layout-nrow: 2
+#| layout-ncol: 2
+
+cleaned_data_cat |>
+  ggplot(mapping = aes(x = categorized_loss, fill = fire_alarm_system_presence)) +
+  geom_bar() +
+  labs(x = "Financial loss group", y = "Number of observations", fill = "Fire alarm presence") +
+  theme(legend.position = "bottom")
+
+cleaned_data_cat |>
+  ggplot(mapping = aes(x = categorized_loss, fill = smoke_alarm_at_fire_origin)) +
+  geom_bar() +
+  labs(x = "Financial loss group", y = "Number of observations", fill = "Smoke alarm presence") +
+  theme(legend.position = "bottom")
+
+cleaned_data_cat |>
+  ggplot(mapping = aes(x = categorized_loss, fill = sprinkler_system_operation)) +
+  geom_bar() +
+  labs(x = "Financial loss group", y = "Number of observations", fill = "Sprinkler system presence") +
+  theme(legend.position = "bottom")
+```
+
+@fig-financial-loss-group shows the distribution of financial loss in group and the presence of fire prevention system.
+In @fig-financial-loss-group-1, the graph shows that incidents with no fire alarm system (red bars) is less than incidents with fire alarm system (blue bars).
+Among the incidents with higher financial loss, there are more cases without fire alarm system.
+In @fig-financial-loss-group-2, Most of the incidents happen with the smoke alarm present.
+Among the incidents with lower financial loss, more smoke alarms are present but not operated compared to higher financial loss cases.
+In @fig-financial-loss-group-3, we can tell that a large proportion of incidents does not have a sprinkler system installed.
+Among those incidents with sprinkler system present, most of them are not operated.
 

### 2024-09-21 paper/paper.qmd
update data overview

@@
 The cleaned data are divided into two groups in accordance to the two most important factors in urban fire incidents: responsiveness of fire department and fire protection equipments, as indicated by @HariMurti2023 and @MasoodRafi2012.
-The first group focuses on the fire services' response time and the loss of life or money, and the second group focuses on the loss of money, reason of fire incidents and the presence or operation of fire protection equipments.
+The first group focuses on the fire services' response time， extent of fire and the loss of money, and the second group focuses on the loss of money, reason of fire incidents and the presence or operation of fire protection equipments.
 All other data features are ignored during the data cleaning process.
@@
 "Estimated Loss in Dollar" is the estimated loss measured in dollar.
+"Extent of Fire" is a categorical indicator from "1" to "11" according to the seriousness of the incident.
 

### 2024-09-22 paper/paper.qmd
add discussion part 1

@@
 
+# Discussion
+
+The analysis of Toronto’s fire incident data reveals critical insights into both the responsiveness of the Toronto Fire Service (TFS) and the role of fire protection systems in mitigating financial losses.
+
+## Toronto Fire Service Responsiveness
+
+The TFS's response times remain generally efficient, with an average time of approximately 4.9 minutes, as reflected in the summary statistics (@tbl-clean-tfs-summary-1).
+The response time fluctuated between 5 to 10 minutes over the 2013–2023 period, with some significant outliers, according to @fig-clean-tfs-diff-dollar-time-1.
+This suggests that while TFS generally responds swiftly, occasional delays may occur due to external factors.
+The response time does not lead to more financial loss, because this correlation is weak, as indicated by @fig-response-vs-money.
+This weak correlation suggests that it does not matter a lot for the Toronto Fire Service to react to fire incident very quickly.
+Other factors, such as the extent of the fire and fire protection measures in place, play a more critical role in determining the financial impact of fire incidents.
+
+The financial loss associated with fire incidents shows a wide range, with most incidents resulting in minimal losses, as shown in @fig-clean-tfs-diff-dollar-time-2.
+However, some outliers, particularly between 2015 and 2024, saw losses exceeding $2 million.
+This suggests that there are rare instances where larger fires occur occasionally, resulting in substantial property damage.
+
+## Impact of Fire Protection Equipment
+
 \newpage

### 2024-09-22 paper/paper.tex
add discussion part 1

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
\PassOptionsToPackage{dvipsnames,svgnames,x11names}{xcolor}
%
\documentclass[
  letterpaper,
  DIV=11,
  numbers=noendperiod]{scrartcl}

\usepackage{amsmath,amssymb}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
\usepackage{lmodern}
\ifPDFTeX\else  
    % xetex/luatex font selection
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\setlength{\emergencystretch}{3em} % prevent overfull lines
\setcounter{secnumdepth}{5}
% Make \paragraph and \subparagraph free-standing
\makeatletter
\ifx\paragraph\undefined\else
  \let\oldparagraph\paragraph
  \renewcommand{\paragraph}{
    \@ifstar
      \xxxParagraphStar
      \xxxParagraphNoStar
  }
  \newcommand{\xxxParagraphStar}[1]{\oldparagraph*{#1}\mbox{}}
  \newcommand{\xxxParagraphNoStar}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
  \let\oldsubparagraph\subparagraph
  \renewcommand{\subparagraph}{
    \@ifstar
      \xxxSubParagraphStar
      \xxxSubParagraphNoStar
  }
  \newcommand{\xxxSubParagraphStar}[1]{\oldsubparagraph*{#1}\mbox{}}
  \newcommand{\xxxSubParagraphNoStar}[1]{\oldsubparagraph{#1}\mbox{}}
\fi
\makeatother

\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}\usepackage{longtable,booktabs,array}
\usepackage{calc} % for calculating minipage widths
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
% definitions for citeproc citations
\NewDocumentCommand\citeproctext{}{}
\NewDocumentCommand\citeproc{mm}{%
  \begingroup\def\citeproctext{#2}\cite{#1}\endgroup}
\makeatletter
 % allow citations to break across lines
 \let\@cite@ofmt\@firstofone
 % avoid brackets around text for \cite:
 \def\@biblabel#1{}
 \def\@cite#1#2{{#1\if@tempswa , #2\fi}}
\makeatother
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newenvironment{CSLReferences}[2] % #1 hanging-indent, #2 entry-spacing
 {\begin{list}{}{%
  \setlength{\itemindent}{0pt}
  \setlength{\leftmargin}{0pt}
  \setlength{\parsep}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1
   \setlength{\leftmargin}{\cslhangindent}
   \setlength{\itemindent}{-1\cslhangindent}
  \fi
  % set entry spacing
  \setlength{\itemsep}{#2\baselineskip}}}
 {\end{list}}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{\hfill\break\parbox[t]{\linewidth}{\strut\ignorespaces#1\strut}}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{\strut#1\strut}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{\strut#1\strut}}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}

\KOMAoption{captions}{tableheading}
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother

\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\usepackage{bookmark}

\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={My title},
  pdfauthor={Lexun Yu},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

\title{My title\thanks{Code and data are available at:
\url{https://github.com/yulexun/toronto-fire}.}}
\usepackage{etoolbox}
\makeatletter
\providecommand{\subtitle}[1]{% add subtitle to \maketitle
  \apptocmd{\@title}{\par {\large #1 \par}}{}{}
}
\makeatother
\subtitle{My subtitle if needed}
\author{Lexun Yu}
\date{September 22, 2024}

\begin{document}
\maketitle
\begin{abstract}
First sentence. Second sentence. Third sentence. Fourth sentence.
\end{abstract}

\renewcommand*\contentsname{Table of contents}
{
\hypersetup{linkcolor=}
\setcounter{tocdepth}{3}
\tableofcontents
}

\section{Introduction}\label{sec-intro}

The urban fire hazard is one of the most pressing issues in this
context, especially in Canada where cities are dealing with such issues
like climate, facilities or population density.As of July 1 2023, the
population in urban area in Canada reached 33,812,133 (Statistics Canada
2024). Not only do fires in highly populated regions result in heavy
losses in terms of property, but also, in the contest of people and the
environment, the consequences are enormous. Also, urban fires are
resources-dependent and require attention from city services and
emergency services, thus indicating the need for prevention, action and
planning based on risk assessment. With such issues in mind, the
knowledge of urban fire hazards in Canadian cities is needed for
creating policies that address public safety and increase urban
resilience.

In Canada, articles about fire incidents has a focus on wildfire. For
instance, Goemans and Ballamingie (2012) discuss the fire mitigation
plan during the 2003 wildfire at Kelowna, British Columbia, while Mamuji
and Rozdilsky (2018) talk about the evacuation during the Fort McMurray
wildfire in Alberta. The researches conducted about urban fire incidents
are done in other parts of the world such as East Asia. Masood Rafi,
Wasiuddin, and Hameed Siddiqui (2012) research the nature and level of
this threat. They conclude that the lack of training in fire department,
shortage of facilities and infrastructure the major issues in Pakistan.
The research by Hari Murti et al. (2023) in Semarang City also emphasize
the importance of community understanding and the installation of fire
protection facilities. This article uses the data provided by
opendatatoronto library (Gelfand 2022) in order to analyze fire
occurrences in the city of Toronto, which is an important research gap
in the study of fire incidents in America. This study seeks to provide a
deeper understanding of fire patterns to improve fire prevention and
emergency response strategies.

In this paper we visualize Toronto's Fire Incidents data.

\section{Data}\label{data}

\subsection{Overview}\label{sec-data-overview}

The data used in this paper is obtained from the opendatatoronto library
(Gelfand 2022). The dataset used is Fire Incidents. According to Gelfand
(2022), it includes fire incidents as defined by the Ontario Fire
Marshal (OFM) up to December 31, 2023. The data gathering and analysis
is done in R (R Core Team 2024) with the following packages:
opendatatoronto (Gelfand 2022), knitr (Xie 2014), tidyverse (Wickham et
al. 2019), ggplot2 (Wickham 2016), dplyr (Wickham et al. 2023), and
lubridate (Grolemund and Wickham 2011).

The cleaned data are divided into two groups in accordance to the two
most important factors in urban fire incidents: responsiveness of fire
department and fire protection equipments, as indicated by Hari Murti et
al. (2023) and Masood Rafi, Wasiuddin, and Hameed Siddiqui (2012). The
first group focuses on the fire services' response time， extent of fire
and the loss of money, and the second group focuses on the loss of
money, reason of fire incidents and the presence or operation of fire
protection equipments. All other data features are ignored during the
data cleaning process.

\subsection{Toronto Fire Service
Responsiveness}\label{toronto-fire-service-responsiveness}

The first group of data shows the responsiveness of Toronto Fire
Services. An example of this dataset is presented in
Table~\ref{tbl-clean-tfs}. ``Alarm Time'' is the time when TFS are
notified of the incident. ``TFS Arrival time'' is the timestamp of first
arriving unit. The difference in minute is calculated at the data
cleaning step as ``TFS Response Time''. ``Estimated Loss in Dollar'' is
the estimated loss measured in dollar. ``Extent of Fire'' is a
categorical indicator from ``1'' to ``11'' according to the seriousness
of the incident.

\begin{longtable}[]{@{}
  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.2041}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.2041}}
  >{\raggedleft\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.1837}}
  >{\raggedleft\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.2551}}
  >{\raggedleft\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.1531}}@{}}

\caption{\label{tbl-clean-tfs}Top rows of cleaned Toronto Fire Service
response time and loss data}

\tabularnewline

\toprule\noalign{}
\begin{minipage}[b]{\linewidth}\raggedright
Alarm Time
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
TFS Arrival Time
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
TFS Response Time
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
Estimated Loss in Dollar
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
Extent of Fire
\end{minipage} \\
\midrule\noalign{}
\endhead
\bottomrule\noalign{}
\endlastfoot
2018-02-25 15:48:34 & 2018-02-25 15:52:04 & 3.500000 & 5000 & 1 \\
2018-02-26 18:11:59 & 2018-02-26 18:15:50 & 3.850000 & 500 & 1 \\
2018-03-03 09:49:14 & 2018-03-03 09:53:09 & 3.916667 & 0 & 1 \\
2018-03-03 17:54:38 & 2018-03-03 17:59:42 & 5.066667 & 15000 & 2 \\
2018-03-03 18:34:35 & 2018-03-03 18:40:47 & 6.200000 & 0 & 1 \\

\end{longtable}

Figure~\ref{fig-clean-tfs-diff-dollar-time-1} shows the line plot of
Toronto Fire Service's response time over the period between 2013 to
2023. The response time fluctuated around a relatively consistent level,
typically between 5 and 10 minutes, with occasional spikes where the
response time increased significantly above the usual range,
particularly around 2014 and 2019.
Figure~\ref{fig-clean-tfs-diff-dollar-time-2} shows the line plot of
estimated dollar loss of the fire incidents. Most of the incidents
resulted in relatively low dollar losses, but there are several spikes
that represent incidents where the loss was significantly higher. Some
of the most notable spikes occurred between 2015 and 2019, with losses
exceeding 2 million dollars.

\begin{figure}

\begin{minipage}{0.50\linewidth}

\centering{

\includegraphics{paper_files/figure-pdf/fig-clean-tfs-diff-dollar-time-1.pdf}

}

\subcaption{\label{fig-clean-tfs-diff-dollar-time-1}TFS response time
over time}

\end{minipage}%
%
\begin{minipage}{0.50\linewidth}

\centering{

\includegraphics{paper_files/figure-pdf/fig-clean-tfs-diff-dollar-time-2.pdf}

}

\subcaption{\label{fig-clean-tfs-diff-dollar-time-2}Estimated dollar
loss over time}

\end{minipage}%

\caption{\label{fig-clean-tfs-diff-dollar-time}TFS response time and
estimated dollar loss over time}

\end{figure}%

Table~\ref{tbl-clean-tfs-summary-1} indicates the summary statistic and
Figure~\ref{fig-clean-tfs-boxplot-1} graphs the boxplot of TFS's
response time. The value mainly range from 4.03 minutes to 5.7 minutes
as shown in Table~\ref{tbl-clean-tfs-summary-1}, with a mean value of
4.911868 minutes. Notably, in Figure~\ref{fig-clean-tfs-boxplot-1},
several outliers are indicated by dots above the upper whisker,
suggesting occasional instances of significantly higher response times.

Table~\ref{tbl-clean-tfs-summary-2} indicates the summary statistic and
Figure~\ref{fig-clean-tfs-boxplot-2} depicts the distribution of
financial losses in boxplot. In Table~\ref{tbl-clean-tfs-summary-2},
values span from 0 dollar to over 3000000 dollars, with median value of
3000 dollars, mean value of 24551.64 dollars and standard deviation of
93520.08 dollars, indicating a skewed distribution.
Figure~\ref{fig-clean-tfs-boxplot-2} reveals numerous outliers above the
upper whisker, highlighting instances of exceptionally high financial
losses.

\begin{figure}

\begin{minipage}{0.50\linewidth}

\centering{

\includegraphics{paper_files/figure-pdf/fig-clean-tfs-boxplot-1.pdf}

}

\subcaption{\label{fig-clean-tfs-boxplot-1}Boxplot of TFS response time}

\end{minipage}%
%
\begin{minipage}{0.50\linewidth}

\centering{

\includegraphics{paper_files/figure-pdf/fig-clean-tfs-boxplot-2.pdf}

}

\subcaption{\label{fig-clean-tfs-boxplot-2}Boxplot of estimated dollar
loss}

\end{minipage}%

\caption{\label{fig-clean-tfs-boxplot}Boxplot of TFS response time data
and estimated dollar loss}

\end{figure}%

\begin{table}

\caption{\label{tbl-clean-tfs-summary}Summary Statistic of TFS response
time data and estimated dollar loss}

\begin{minipage}{0.50\linewidth}

\subcaption{\label{tbl-clean-tfs-summary-1}Summary of TFS response time}

\centering{

\begin{tabular}{lr}
\toprule
Statistic & Value\\
\midrule
Min. & 0.300000\\
1st Qu. & 4.033333\\
Median & 4.816667\\
Mean & 4.911166\\
3rd Qu. & 5.716667\\
Max. & 24.433333\\
SD & 1.399413\\
\bottomrule
\end{tabular}

}

\end{minipage}%
%
\begin{minipage}{0.50\linewidth}

\subcaption{\label{tbl-clean-tfs-summary-2}Summary of estimated dollar loss}

\centering{

\begin{tabular}{lr}
\toprule
Statistic & Value\\
\midrule
Min. & 0.00\\
1st Qu. & 500.00\\
Median & 3000.00\\
Mean & 26584.26\\
3rd Qu. & 15000.00\\
Max. & 3000000.00\\
SD & 104422.21\\
\bottomrule
\end{tabular}

}

\end{minipage}%

\end{table}%

Figure~\ref{fig-bar-extent} illustrates that most of the observed fires
are relatively small in size. The x-axis depicts the extent of the fire,
while the y-axis shows the frequency of observations. A notably tall bar
at the start signifies a high occurrence of smaller fires. As fire size
increases along the x-axis, the number of observations drops sharply,
indicating that larger fires are far less frequent.

\begin{figure}

\centering{

\includegraphics{paper_files/figure-pdf/fig-bar-extent-1.pdf}

}

\caption{\label{fig-bar-extent}Distribution of extent of fire}

\end{figure}%

Figure~\ref{fig-response-vs-money} shows a line of best fit, which
indicates a weak correlation between TFS response time and estimated
dollar loss. There is a wide range of TFS response times for lower
dollar losses.

\begin{figure}

\centering{

\includegraphics{paper_files/figure-pdf/fig-response-vs-money-1.pdf}

}

\caption{\label{fig-response-vs-money}Line of best fit between TFS
response time and dollar loss}

\end{figure}%

\subsection{Fire Protection
Equipments}\label{fire-protection-equipments}

The second group of data shows the money loss, cause of fire and the
presence of fire prevention equipments. Table~\ref{tbl-clean-cat} shows
the first five rows of the second group of data. ``Estimated Loss in
Dollar'' is the estimated loss measured in dollar. ``Area of origin''
indicates the area fire originate. ``Ignition Source'' shows the object
causing fire. ``Fire Alarm Status'', ``Smoke Alarm Status'' and
``Sprinkler System Status'' indicates the presence and operation of fire
protection equipments. In the second group of data, `PO' = System
present, `N' = System not present, and `P' = System present but not
operated.

\begin{longtable}[]{@{}
  >{\raggedleft\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.2137}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.1282}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.1368}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.1538}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.1624}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.2051}}@{}}

\caption{\label{tbl-clean-cat}Top rows of cleaned data showing Area of
Origin, Ignition source and Fire, Smoke, Sprnkler System Presence.}

\tabularnewline

\toprule\noalign{}
\begin{minipage}[b]{\linewidth}\raggedleft
Estimated Loss in Dollar
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Area of Origin
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Ignition Source
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Fire Alarm Status
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Smoke Alarm Status
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Sprinkler System Status
\end{minipage} \\
\midrule\noalign{}
\endhead
\bottomrule\noalign{}
\endlastfoot
5000 & 28 & 41 & P & PO & P \\
500 & 24 & 11 & P & PO & N \\
0 & 24 & 11 & N & PO & P \\
15000 & 25 & 24 & N & N & N \\
0 & 24 & 11 & P & PO & P \\

\end{longtable}

Figure~\ref{fig-origin} in \hyperref[sec-origin-fig]{Appendix A.1}
illustrates the distribution of the fire incidents' area of origin. It
consists 71 unique categories. The most common area of origin is
category 24, which is cooking area or kitchen. Other common areas
include category 64: Porch or Balcony, 22: Sleeping Area or Bedroom, 21:
Living Area, and 27: Laundry Area. However, fires in these locations
occur significantly less often compared to those in kitchens.

Figure~\ref{fig-source} in \hyperref[sec-source-fig]{Appendix A.2} shows
the distribution of ignition source in fire incidents. It contains 73
unique categories. The most common ignition source is 11: Stove,
Range-top burner and 71: Smoker's Articles. Other common ignition
sources include 55: Candle, 43: Clothes Dryer, and 12: Oven.

\begin{figure}

\begin{minipage}{0.50\linewidth}

\centering{

\includegraphics{paper_files/figure-pdf/fig-financial-loss-group-1.pdf}

}

\subcaption{\label{fig-financial-loss-group-1}Distribution of financial
loss group, and fire alarm presence}

\end{minipage}%
%
\begin{minipage}{0.50\linewidth}

\centering{

\includegraphics{paper_files/figure-pdf/fig-financial-loss-group-2.pdf}

}

\subcaption{\label{fig-financial-loss-group-2}Distribution of financial
loss group, and smoke alarm presence}

\end{minipage}%
\newline
\begin{minipage}{0.50\linewidth}

\centering{

\includegraphics{paper_files/figure-pdf/fig-financial-loss-group-3.pdf}

}

\subcaption{\label{fig-financial-loss-group-3}Distribution of financial
loss group, and sprinkler system presence}

\end{minipage}%

\caption{\label{fig-financial-loss-group}Distribution of financial loss
group, and fire prevention system presence}

\end{figure}%

Figure~\ref{fig-financial-loss-group} shows the distribution of
financial loss in group and the presence of fire prevention system. In
Figure~\ref{fig-financial-loss-group-1}, the graph shows that incidents
with no fire alarm system (red bars) is less than incidents with fire
alarm system (blue bars). Among the incidents with higher financial
loss, there are more cases without fire alarm system. In
Figure~\ref{fig-financial-loss-group-2}, Most of the incidents happen
with the smoke alarm present. Among the incidents with lower financial
loss, more smoke alarms are present but not operated compared to higher
financial loss cases. In Figure~\ref{fig-financial-loss-group-3}, we can
tell that a large proportion of incidents does not have a sprinkler
system installed. Among those incidents with sprinkler system present,
most of them are not operated.

\section{Discussion}\label{discussion}

The analysis of Toronto's fire incident data reveals critical insights
into both the responsiveness of the Toronto Fire Service (TFS) and the
role of fire protection systems in mitigating financial losses.

\subsection{Toronto Fire Service
Responsiveness}\label{toronto-fire-service-responsiveness-1}

The TFS's response times remain generally efficient, with an average
time of approximately 4.9 minutes, as reflected in the summary
statistics (Table~\ref{tbl-clean-tfs-summary-1}). The response time
fluctuated between 5 to 10 minutes over the 2013--2023 period, with some
significant outliers, according to
Figure~\ref{fig-clean-tfs-diff-dollar-time-1}. This suggests that while
TFS generally responds swiftly, occasional delays may occur due to
external factors. The response time does not lead to more financial
loss, because this correlation is weak, as indicated by
Figure~\ref{fig-response-vs-money}. This weak correlation suggests that
it does not matter a lot for the Toronto Fire Service to react to fire
incident very quickly. Other factors, such as the extent of the fire and
fire protection measures in place, play a more critical role in
determining the financial impact of fire incidents.

The financial loss associated with fire incidents shows a wide range,
with most incidents resulting in minimal losses, as shown in
Figure~\ref{fig-clean-tfs-diff-dollar-time-2}. However, some outliers,
particularly between 2015 and 2024, saw losses exceeding \$2 million.
This suggests that there are rare instances where larger fires occur
occasionally, resulting in substantial property damage.

\subsection{Impact of Fire Protection
Equipment}\label{impact-of-fire-protection-equipment}

\newpage

\appendix

\section{Appendix}\label{sec-appendix}

\subsection{Graph of areas of origin of fire incidents by number of
occurrences}\label{sec-origin-fig}

\begin{verbatim}
Index of 'Area of Origin':
\end{verbatim}

\begin{verbatim}
11 - Lobby, Entranceway
12 - Hallway, Corridor
13 - Stairway, Escalator
18 - Covered Court, Atrium, mall concourse
19 - Other Means of Egress
21 - Living Area (e.g. living, TV, recreation, etc)
22 - Sleeping Area or Bedroom (inc. patients room, dormitory, etc)
23 - Dining or Beverage Area (inc mess, canteen, lunchroom, cafeteria
24 - Cooking Area or Kitchen
25 - Washroom or Bathroom (toilet,restroom/locker room)
26 - Sauna
27 - Laundry Area
28 - Office
29 - Electronic Equipment
30 - Sales, Showroom Area
31 - Process Manufacturing (inc manf, prod assembly, repair)
32 - Assembly Area (inc school room,spectator area, church, etc)
33 - Laboratory
34 - Operating Room, Treatment or Examination Area
35 - Performance Area (inc stage, rink, boxing ring, gym floor, altar
36 - Backstage, dressing room
39 - Other Functional Area
41 - Closet (eg. clothes, broom, linen closet, etc.)
42 - Garage
43 - Locker (apartment storage)
44 - Trash, Rubbish Storage (inc garbage chute room, garbage/industri
45 - Supply Storage Room (inc maintenance/office/document storage, et
46 - Product Storage (inc products or materials awaiting manuf, assembly)
47 - Shipping/Receiving/Loading Platform
48 - Records storage area (inc vaults)
49 - Other Storage Area
50 - Basement/cellar (not partitioned)
51 - Elevator (includes shaft)
52 - HVAC Equipment Room (furnace room, water heater closet, boiler)
53 - Chimney/Flue Pipe
54 - Incinerator Room
55 - Mechanical/Electrical Services Room
56 - Conveyor Shaft or Chute (inc dumbwaiter, laundry chute, garbage
57 - Ducting - Heating, Air Conditioning
58 - Ducting - Exhaust (inc cooking, fumes, etc.)
59 - Utility Shaft (eg. electrical wiring/phone, etc.)
60 - Other Building Services/Support Facilities
61 - Exterior Wall
62 - Roof
63 - Awning or Canopy
64 - Porch or Balcony
65 - Crawl Space (includes sub-structure)
66 - Concealed Ceiling Area
67 - Concealed Floor Area
68 - Concealed Wall Area
69 - Attic Area
70 - Other Structural Area
71 - Open Area (inc lawn, field, farmyard, park, playing field, pier,
72 - Court, Patio, Terrace
73 - Parking Area, Parking Lot
74 - Storage Area (outside)
75 - Trash, rubbish area (outside)
78 - Attached Deck
79 - Other Outside Area
81 - Engine Area
82 - Running Gear (inc wheels and braking systems, transmission syste
83 - Electrical Systems
84 - Fuel Systems (eg. fuel tank, etc.)
85 - Operator/Control Area
86 - Passenger Area
87 - Trunk/Cargo Area
89 - Other Vehicle Area
91 - Multiple Areas of Origin
92 - Residential/Business: Restaurant area
93 - Residential/Business: Other busines area
\end{verbatim}

\begin{figure}

\centering{

\includegraphics{paper_files/figure-pdf/fig-origin-1.pdf}

}

\caption{\label{fig-origin}Areas of origin of fire incidents by number
of occurrences}

\end{figure}%

\newpage

\subsection{Graph of ignition source of fire incidents by number of
occurrences}\label{sec-source-fig}

\begin{verbatim}
Index of 'Ignition Source':
\end{verbatim}

\begin{verbatim}
100 - Outdoor fireplace/heater
101 - Exposure, source structure detached
102 - Exposure, source structure semi-detached or attached
103 - Exposure, source outside storage container, tank
104 - Exposure, source open fire (inc campfire, rubbish fire)
106 - Exposure, source grass, shrubs, trees
107 - Exposure, source vehicle (outside structure)
108 - Exposure, source other
11 - Stove, Range-top burner
12 - Oven
13 - Microwave
14 - Open Fired Barbeque - Fixed or Portable
15 - Range Hood
16 - Deep Fat Fryer
17 - Wood burning stove
19 - Other Cooking Items (eg Toaster, Kettle, elec frying pan)
20 - Service/Utility Lines (includes power/hydro transmission lines)
21 - Transformer
22 - Meter
23 - Distribution Equipment (includes panel boards, fuses, circuit br
24 - Circuit Wiring - Copper
25 - Circuit Wiring - Aluminum
26 - Terminations-Copper (incl receptacles, switches, lights)
27 - Terminations-Aluminum (incl receptables, switches, lights)
28 - Cord, Cable for Appliance, Electrical Articles
29 - Extension Cord, Temporary Wiring
30 - Other Electrical Distribution Item
31 - Central Heating/Cooling Unit
32 - Water Heater
33 - Space Heater - Fixed
34 - Space Heater - Portable
35 - Fireplace - Factory Built
36 - Fireplace - Masonry
37 - Fireplace Insert
38 - Chimney - Factory Built
39 - Chimney - Masonry
40 - Flue Pipe
41 - Other Heating Equipment
42 - Television, Radio, Stereo, Tape Recorder, etc.
43 - Clothes Dryer
44 - Iron, Pressing Machine
45 - Washing Machine
46 - Electric Blanket, Heating Pad
47 - Refrigerator, Freezer (includes vending machine)
48 - Air Conditioner - Window or Room Unit
49 - Other Appliances
51 - Incandescent Lamp - Light Bulb, Spotlight
52 - Florescent Lamp (includes ballast)
53 - Christmas Lights, Decorative Lighting
54 - Lamp (eg. coal, oil, naphtha, etc.)
55 - Candle
56 - Halogen Lamp or light
59 - Other Lighting Equipment
61 - Incinerator
62 - Heat Treatment Equipment (eg. furnace, oven, kiln, quench tanks,
63 - Painting Equipment
64 - Chemical Processing Equipment (eg. reactors, distilling units, e
69 - Other Processing Equipment
71 - Smoker's Articles (eg. cigarettes, cigars, pipes already ignited
72 - Cutting/Welding Equipment
73 - Blow Torch, Bunsen Burner
74 - Salamander
75 - Matches (open flame)
76 - Lighters (open flame)
77 - Matches or Lighters (unable to distinguish)
79 - Other Open Flame Tools/Smokers' Articles
80 - Portable generator
81 - Vehicle - Electrical
82 - Vehicle - Mechanical
83 - Other Electrical
84 - Other Mechanical
85 - Vehicle collision
88 - Multiple Ignition Source or Igniting Equipment (suspected arson)
91 - Fireworks
92 - Open Fire (eg. camp fire, rubbish fire, etc.)
93 - Hot Ashes, Embers, Spark
94 - Static Electricity (spark)
95 - Lightning
96 - Chemical Reaction (eg. spontaneous combustion, etc.)
97 - Rekindle
\end{verbatim}

\begin{figure}

\centering{

\includegraphics{paper_files/figure-pdf/fig-source-1.pdf}

}

\caption{\label{fig-source}Ignition Source of fire incidents by number
of occurrences}

\end{figure}%

\subsection{Attribution Statement}\label{attribution-statement}

Contains information licensed under the Open Government Licence --
Toronto. Visit: \url{https://open.toronto.ca/open-data-license/}

\newpage

\section*{Bibliography}\label{bibliography}
\addcontentsline{toc}{section}{Bibliography}

\phantomsection\label{refs}
\begin{CSLReferences}{1}{0}
\bibitem[\citeproctext]{ref-opendatatoronto}
Gelfand, Sharla. 2022. \emph{Opendatatoronto: Access the City of Toronto
Open Data Portal}.
\url{https://CRAN.R-project.org/package=opendatatoronto}.

\bibitem[\citeproctext]{ref-Goemans2012}
Goemans, Magdalene, and Patricia Ballamingie. 2012. {``Forest as Hazard,
Forest as Victim: Community Perspectives and Disaster Mitigation in the
Aftermath of Kelowna's 2003 Wildfires.''} \emph{Canadian Geographies /
Géographies Canadiennes} 57 (1): 56--71.
\url{https://doi.org/10.1111/j.1541-0064.2012.00447.x}.

\bibitem[\citeproctext]{ref-lubridate}
Grolemund, Garrett, and Hadley Wickham. 2011. {``Dates and Times Made
Easy with {lubridate}.''} \emph{Journal of Statistical Software} 40 (3):
1--25. \url{https://www.jstatsoft.org/v40/i03/}.

\bibitem[\citeproctext]{ref-HariMurti2023}
Hari Murti, Raditya, Hendra Adi Wijaya, Indira Laksmi Widuri, Julmadian
Abda, Mada Sophianingrum, Muhammad Rizki Islami, Ahady Farrel
Febriyanto, and Eduardo Erlangga Drestanta. 2023. {``Risk Assessment of
Fire Hazards in Semarang City Residential Areas.''} \emph{Jurnal Teknik
Sipil Dan Perencanaan} 25 (1): 52--61.
\url{https://doi.org/10.15294/jtsp.v25i1.42955}.

\bibitem[\citeproctext]{ref-Mamuji2018}
Mamuji, Aaida A., and Jack L. Rozdilsky. 2018. {``Wildfire as an
Increasingly Common Natural Disaster Facing Canada: Understanding the
2016 Fort McMurray Wildfire.''} \emph{Natural Hazards} 98 (1): 163--80.
\url{https://doi.org/10.1007/s11069-018-3488-4}.

\bibitem[\citeproctext]{ref-MasoodRafi2012}
Masood Rafi, Muhammad, Syed Wasiuddin, and Salman Hameed Siddiqui. 2012.
{``Assessment of Fire Hazard in Pakistan.''} \emph{Disaster Prevention
and Management: An International Journal} 21 (1): 71--84.
\url{https://doi.org/10.1108/09653561211202719}.

\bibitem[\citeproctext]{ref-citeR}
R Core Team. 2024. \emph{R: A Language and Environment for Statistical
Computing}. Vienna, Austria: R Foundation for Statistical Computing.
\url{https://www.R-project.org/}.

\bibitem[\citeproctext]{ref-statsCanada}
Statistics Canada. 2024. {``Components of Population Change by Census
Metropolitan Area and Census Agglomeration, 2021 Boundaries.''}
Government of Canada. \url{https://doi.org/10.25318/1710014901-ENG}.

\bibitem[\citeproctext]{ref-ggplot2}
Wickham, Hadley. 2016. \emph{Ggplot2: Elegant Graphics for Data
Analysis}. Springer-Verlag New York.
\url{https://ggplot2.tidyverse.org}.

\bibitem[\citeproctext]{ref-tidyverse}
Wickham, Hadley, Mara Averick, Jennifer Bryan, Winston Chang, Lucy
D'Agostino McGowan, Romain François, Garrett Grolemund, et al. 2019.
{``Welcome to the {tidyverse}.''} \emph{Journal of Open Source Software}
4 (43): 1686. \url{https://doi.org/10.21105/joss.01686}.

\bibitem[\citeproctext]{ref-dplyr}
Wickham, Hadley, Romain François, Lionel Henry, Kirill Müller, and Davis
Vaughan. 2023. \emph{Dplyr: A Grammar of Data Manipulation}.
\url{https://CRAN.R-project.org/package=dplyr}.

\bibitem[\citeproctext]{ref-knitr}
Xie, Yihui. 2014. {``Knitr: A Comprehensive Tool for Reproducible
Research in {R}.''} In \emph{Implementing Reproducible Computational
Research}, edited by Victoria Stodden, Friedrich Leisch, and Roger D.
Peng. Chapman; Hall/CRC.

\end{CSLReferences}

\end{document}

### 2024-09-22 paper/paper.qmd
update paper

--- 2024-09-21 paper/paper.qmd
@@
 
+# Discussion
+
+The analysis of Toronto’s fire incident data reveals critical insights into both the responsiveness of the Toronto Fire Service (TFS) and the role of fire protection systems in mitigating financial losses.
+
+## Toronto Fire Service Responsiveness
+
+The Toronto Fire Service (TFS) generally has efficient response times, averaging about 4.9 minutes, as noted in the summary statistics (@tbl-clean-tfs-summary-1).
+Over the period from 2013 to 2023, response times varied between 5 and 10 minutes, with some notable outliers, according to @fig-clean-tfs-diff-dollar-time-1.
+This indicates that while TFS usually responds quickly, there can be occasional delays due to outside factors.
+Importantly, these response times do not lead to significant financial losses, because the relationship between response time and financial impact is weak, as indicated by @fig-response-vs-money.
+This implies that for the TFS, rapid response to fire incidents is not the most critical factor.
+Other elements, such as the severity of the fire and existing fire protection measures, are more decisive in determining the financial consequences of fire events.
+
+Financial losses from fire incidents vary greatly, with most cases resulting in low losses, as illustrated in @fig-clean-tfs-diff-dollar-time-2.
+However, certain outliers, especially from 2015 to 2024, have reported losses exceeding $2 million.
+This indicates that larger fires occur occasionally, leading to considerable property damage.
+
+## Impact of Fire Protection Equipment
+The presence and operation of fire protection equipment, particularly alarms and sprinklers, had a noticeable effect on financial losses.
+According to @fig-financial-loss-group, incidents where fire alarms were absent or failed to operate resulted in higher financial losses.
+This is consistent with previous researched carried out including the one conducted by @HariMurti2023 which further corroborates the studies that the absence of effective fire prevention systems makes the aftermath of fire destruction more damaging.
+SAdditionally, the data reveal that while smoke alarms were present in most cases, sprinklers were often absent or not operational.
+This underlines the importance of incorporating sprinkler systems particularly in these areas where most fire outbreaks are recorded such as kitchens in accordance with statistics.
+
+The results also indicate that there are other factors such as the location of the fire’s ignition and its sources that are critical in determining the outcome.
+According to @fig-origin and @fig-source, kitchens and living rooms are the most common areas where fires start with most cases arising from stovetops and smokers’ articles being the most common.
+Such information makes it evident that fire outbreak should be prevented by targeting those measures in these severe areas.
+For instance fire incidents would be reduced considerable alongside its severity by making people understand the dangers of kitchen fires and the need to manage sprinkler and smoke alarm systems so that they do not malfunction.
+
+## Future Directions and Limitations
+
+This assessment is of great importance in the context of fire prevention and emergency response plans in the city of Toronto.
+First of all, the response times from TFS, apart from being fairly constant, also fluctuate at times, justifying the strategy that more improvements in the fire service will be in order in particular areas where there are delays.
+Furthermore, the data emphasize the great importance of fire protection systems in decreases in losses and the necessity of not only having fire protection systems in place but having them in working order as well.
+
+There are a few limitations on this research.
+First, there are other factors affecting TFS response time, such as local extreme weather or traffic.
+The response time data may not show the full picture of TFS's responsiveness.
+Second, the presence of fire protection equipment is oversimplified in the dataset.
+It is grouped to only three main categories.
+The state of fire alarm, smoke alarm and sprinkler system has to be analyzed more thoughtfully during the data collection process for a more useful analysis.
+
+Overall, this analysis contributes to a deeper understanding of fire risks in urban environments and supports the development of targeted interventions aimed at minimizing fire-related damages.
+
 \newpage
@@
 \newpage
+
 # Bibliography

### 2024-09-22 paper/paper.qmd
final paper

@@
 ---
-title: "My title"
-subtitle: "My subtitle if needed"
+title: "Assessing Fire Response Efficiency and the Crucial Role of Protection Systems in Reducing Damage"
+subtitle: "Insights from Toronto’s Fire Incidentsfinal"
 author: 
@@
 date-format: long
-abstract: "First sentence. Second sentence. Third sentence. Fourth sentence."
+abstract: "This paper analyzes fire incidents in Toronto between 2013 and 2023, focusing on the response times of the Toronto Fire Service (TFS) and the impact of fire protection systems.
+The study finds that TFS usually responds within 5 to 10 minutes, but the correlation between response time and financial losses is weak, and other factors like fire protection equipment play a larger role in mitigating damage.
+Fire alarms and sprinkler systems in present and operational reduce losses significantly.
+This research the importance of maintaining fire protection systems and improving fire service efficiency to enhance urban fire safety."
 format: pdf
@@
 In this paper we visualize Toronto's Fire Incidents data.
+The remainder of the paper is structured as follows.
+[Section 2](#sec-data) discusses the raw and cleaned data, as well as the summary statistic.
+Then in [Section 2.2](#sec-data-tfs) and [Section 2.3](#sec-data-fpe), We discuss the trend and distribution of TFS response time and fire protection equipment presence.
+In [Section 3](#sec-discuss), we provide analysis of the trend and correlation in [Section 2](#sec-data), as well as the limitations and future directions in resolving urban fire incidents.
 
-# Data
+# Data {#sec-data}
 
@@
 
-## Toronto Fire Service Responsiveness
+## Toronto Fire Service Responsiveness {#sec-data-tfs}
 
@@
 
-## Fire Protection Equipments
+## Fire Protection Equipments {#sec-data-fpe}
 
@@
 
-# Discussion
+# Discussion {#sec-discuss}
 
@@
 
-The Toronto Fire Service (TFS) generally has efficient response times, averaging about 4.9 minutes, as noted in the summary statistics (@tbl-clean-tfs-summary-1).
+The Toronto Fire Service (TFS) generally has efficient response times, averaging about 4.9 minutes, as noted in the summary statistics [@tbl-clean-tfs-summary-1].
 Over the period from 2013 to 2023, response times varied between 5 and 10 minutes, with some notable outliers, according to @fig-clean-tfs-diff-dollar-time-1.

### 2024-09-22 paper/paper.qmd
fix title

@@
 title: "Assessing Fire Response Efficiency and the Crucial Role of Protection Systems in Reducing Damage"
-subtitle: "Insights from Toronto’s Fire Incidentsfinal"
+subtitle: "Insights from Toronto’s Fire Incidents"
 author: 

### 2024-09-23 paper/paper.qmd
update paper structure in intro

@@
 The remainder of the paper is structured as follows.
-[Section 2](#sec-data) discusses the raw and cleaned data, as well as the summary statistic.
-Then in [Section 2.2](#sec-data-tfs) and [Section 2.3](#sec-data-fpe), We discuss the trend and distribution of TFS response time and fire protection equipment presence.
-In [Section 3](#sec-discuss), we provide analysis of the trend and correlation in [Section 2](#sec-data), as well as the limitations and future directions in resolving urban fire incidents.
+@sec-data discusses the raw and cleaned data, as well as the summary statistic.
+Then in @sec-data-tfs and @sec-data-fpe, We discuss the trend and distribution of TFS response time and fire protection equipment presence.
+In @sec-discuss, we provide analysis of the trend and correlation in @sec-data, as well as the limitations and future directions in resolving urban fire incidents.
 

### 2024-09-23 paper/paper.qmd
fix grammar error

@@
 # Introduction {#sec-intro}
-The urban fire hazard is one of the most pressing issues in this context, especially in Canada where cities are dealing with such issues like climate, facilities or population density.As of July 1 2023, the population in urban area in Canada reached 33,812,133 [@statsCanada].
+The urban fire hazard is one of the most pressing issues in this context, especially in Canada where cities are dealing with such issues like climate, facilities or population density.
+As of July 1 2023, the population in urban area in Canada reached 33,812,133 [@statsCanada].
 Not only do fires in highly populated regions result in heavy losses in terms of property, but also, in the contest of people and the environment, the consequences are enormous.

### 2024-09-23 paper/paper.qmd
fix grammar

@@
 abstract: "This paper analyzes fire incidents in Toronto between 2013 and 2023, focusing on the response times of the Toronto Fire Service (TFS) and the impact of fire protection systems.
-The study finds that TFS usually responds within 5 to 10 minutes, but the correlation between response time and financial losses is weak, and other factors like fire protection equipment play a larger role in mitigating damage.
+The study finds that TFS usually responds within 5 to 10 minutes, but the correlation between response time and economic losses is weak, and other factors like fire protection equipment play a larger role in mitigating damage.
 Fire alarms and sprinkler systems in present and operational reduce losses significantly.
@@
 # Introduction {#sec-intro}
-The urban fire hazard is one of the most pressing issues in this context, especially in Canada where cities are dealing with such issues like climate, facilities or population density.
-As of July 1 2023, the population in urban area in Canada reached 33,812,133 [@statsCanada].
-Not only do fires in highly populated regions result in heavy losses in terms of property, but also, in the contest of people and the environment, the consequences are enormous.
-Also, urban fires are resources-dependent and require attention from city services and emergency services, thus indicating the need for prevention, action and planning based on risk assessment.
+The urban fire hazard is one of the most pressing issues in this context, especially in Canada where cities are dealing with such issues like climate, facilities, or population density.
+As of July 1, 2023, the population in urban area in Canada reached 33,812,133 [@statsCanada].
+Not only do fires in highly populated regions result in heavy losses in terms of property, but also, in the context of people and the environment, the consequences are enormous.
+Also, urban fires are resource-dependent and require attention from city services and emergency services, thus indicating the need for prevention, action and planning based on risk assessment.
 With such issues in mind, the knowledge of urban fire hazards in Canadian cities is needed for creating policies that address public safety and increase urban resilience.
 
-In Canada, articles about fire incidents has a focus on wildfire.
+In Canada, articles about fire incidents have a focus on wildfire.
 For instance, @Goemans2012 discuss the fire mitigation plan during the 2003 wildfire at Kelowna, British Columbia, while @Mamuji2018 talk about the evacuation during the Fort McMurray wildfire in Alberta.
-The researches conducted about urban fire incidents are done in other parts of the world such as East Asia.
+Research conducted about urban fire incidents is done in other parts of the world such as East Asia.
 @MasoodRafi2012 research the nature and level of this threat.
-They conclude that the lack of training in fire department, shortage of facilities and infrastructure the major issues in Pakistan.
-The research by @HariMurti2023 in Semarang City also emphasize the importance of community understanding and the installation of fire protection facilities.
-This article uses the data provided by opendatatoronto library [@opendatatoronto] in order to analyze fire occurrences in the city of Toronto, which is an important research gap in the study of fire incidents in America.
+They conclude that the lack of training in fire department, shortage of facilities and infrastructure the key issues in Pakistan.
+The research by @HariMurti2023 in Semarang City also emphasizes the importance of community understanding and the installation of fire protection facilities.
+This article uses the data provided by opendatatoronto library [@opendatatoronto] to analyze fire occurrences in the city of Toronto, which is an important research gap in the study of fire incidents in America.
 This study seeks to provide a deeper understanding of fire patterns to improve fire prevention and emergency response strategies.
@@
 @sec-data discusses the raw and cleaned data, as well as the summary statistic.
-Then in @sec-data-tfs and @sec-data-fpe, We discuss the trend and distribution of TFS response time and fire protection equipment presence.
+Then in @sec-data-tfs and @sec-data-fpe, we discuss the trend and distribution of TFS response time and fire protection equipment presence.
 In @sec-discuss, we provide analysis of the trend and correlation in @sec-data, as well as the limitations and future directions in resolving urban fire incidents.
@@
 
-The cleaned data are divided into two groups in accordance to the two most important factors in urban fire incidents: responsiveness of fire department and fire protection equipments, as indicated by @HariMurti2023 and @MasoodRafi2012.
-The first group focuses on the fire services' response time， extent of fire and the loss of money, and the second group focuses on the loss of money, reason of fire incidents and the presence or operation of fire protection equipments.
+The cleaned data are divided into two groups in accordance with the two most crucial factors in urban fire incidents: responsiveness of fire department and fire protection equipment, as indicated by @HariMurti2023 and @MasoodRafi2012.
+The first group focuses on the fire services' response time, extent of fire and the loss of money, and the second group focuses on the loss of money, reason of fire incidents and the presence or operation of fire protection equipment.
 All other data features are ignored during the data cleaning process.
@@
 "Alarm Time" is the time when TFS are notified of the incident.
-"TFS Arrival time" is the timestamp of first arriving unit.
+"TFS Arrival time" is the timestamp of the first arriving unit.
 The difference in minute is calculated at the data cleaning step as "TFS Response Time".
-"Estimated Loss in Dollar" is the estimated loss measured in dollar.
+"Estimated Loss in Dollars" is the estimated loss measured in dollars.
 "Extent of Fire" is a categorical indicator from "1" to "11" according to the seriousness of the incident.
@@
   kable(
-    col.names = c("Alarm Time", "TFS Arrival Time", "TFS Response Time", "Estimated Loss in Dollar", "Extent of Fire"),
+    col.names = c("Alarm Time", "TFS Arrival Time", "TFS Response Time", "Estimated Loss in Dollars", "Extent of Fire"),
       )
@@
 @fig-clean-tfs-diff-dollar-time-1 shows the line plot of Toronto Fire Service's response time over the period between 2013 to 2023.
-The response time fluctuated around a relatively consistent level, typically between 5 and 10 minutes, with occasional spikes where the response time increased significantly above the usual range, particularly around 2014 and 2019.
+The response time fluctuated around a consistent level, typically between 5 and 10 minutes, with occasional spikes where the response time increased significantly above the usual range, particularly around 2014 and 2019.
 @fig-clean-tfs-diff-dollar-time-2 shows the line plot of estimated dollar loss of the fire incidents.
-Most of the incidents resulted in relatively low dollar losses, but there are several spikes that represent incidents where the loss was significantly higher.
+Most of the incidents resulted in low dollar losses, but there are several spikes that represent incidents where the loss was significantly higher.
 Some of the most notable spikes occurred between 2015 and 2019, with losses exceeding 2 million dollars.
@@
 @tbl-clean-tfs-summary-1 indicates the summary statistic and @fig-clean-tfs-boxplot-1 graphs the boxplot of TFS's response time.
-The value mainly range from 4.03 minutes to 5.7 minutes as shown in @tbl-clean-tfs-summary-1, with a mean value of 4.911868 minutes.
+The value ranges from 4.03 minutes to 5.7 minutes as shown in @tbl-clean-tfs-summary-1, with a mean value of 4.911868 minutes.
 Notably, in @fig-clean-tfs-boxplot-1, several outliers are indicated by dots above the upper whisker, suggesting occasional instances of significantly higher response times.
 
-@tbl-clean-tfs-summary-2 indicates the summary statistic and @fig-clean-tfs-boxplot-2 depicts the distribution of financial losses in boxplot.
+@tbl-clean-tfs-summary-2 indicates the summary statistic and @fig-clean-tfs-boxplot-2 depicts the distribution of economic losses in boxplot.
 In @tbl-clean-tfs-summary-2, values span from 0 dollar to over 3000000 dollars, with median value of 3000 dollars, mean value of 24551.64 dollars and standard deviation of 93520.08 dollars, indicating a skewed distribution.
-@fig-clean-tfs-boxplot-2 reveals numerous outliers above the upper whisker, highlighting instances of exceptionally high financial losses.
+@fig-clean-tfs-boxplot-2 reveals numerous outliers above the upper whisker, highlighting instances of exceptionally high economic losses.
 
@@
 
-@fig-bar-extent illustrates that most of the observed fires are relatively small in size.
+@fig-bar-extent illustrates that most of the observed fires are relatively small.
 The x-axis depicts the extent of the fire, while the y-axis shows the frequency of observations.
@@
 
-The second group of data shows the money loss, cause of fire and the presence of fire prevention equipments.
+The second group of data shows the money loss, cause of fire and the presence of fire prevention equipment.
 @tbl-clean-cat shows the first five rows of the second group of data.
-"Estimated Loss in Dollar" is the estimated loss measured in dollar.
-"Area of origin" indicates the area fire originate.
+"Estimated Loss in Dollars" is the estimated loss measured in dollars.
+"Area of origin" indicates the area fire originates.
 "Ignition Source" shows the object causing fire.
-"Fire Alarm Status", "Smoke Alarm Status" and "Sprinkler System Status" indicates the presence and operation of fire protection equipments.
+"Fire Alarm Status", "Smoke Alarm Status" and "Sprinkler System Status" indicate the presence and operation of fire protection equipment.
 In the second group of data, 'PO' = System present, 'N' = System not present, and 'P' = System present but not operated.
@@
   kable(
-    col.names = c("Estimated Loss in Dollar", "Area of Origin", "Ignition Source", "Fire Alarm Status"
+    col.names = c("Estimated Loss in Dollars", "Area of Origin", "Ignition Source", "Fire Alarm Status"
     , "Smoke Alarm Status", "Sprinkler System Status"),
@@
 @fig-origin in [Appendix A.1](#sec-origin-fig) illustrates the distribution of the fire incidents' area of origin.
-It consists 71 unique categories.
+It consists of 71 unique categories.
 The most common area of origin is category 24, which is cooking area or kitchen.
@@
 #| label: fig-financial-loss-group
-#| fig-cap: "Distribution of financial loss group, and fire prevention system presence"
-#| fig-subcap: ["Distribution of financial loss group, and fire alarm presence", "Distribution of financial loss group, and smoke alarm presence", "Distribution of financial loss group, and sprinkler system presence"]
+#| fig-cap: "Distribution of economic loss group, and fire prevention system presence"
+#| fig-subcap: ["Distribution of economic loss group, and fire alarm presence", "Distribution of economic loss group, and smoke alarm presence", "Distribution of economic loss group, and sprinkler system presence"]
 #| echo: false
@@
 
-@fig-financial-loss-group shows the distribution of financial loss in group and the presence of fire prevention system.
-In @fig-financial-loss-group-1, the graph shows that incidents with no fire alarm system (red bars) is less than incidents with fire alarm system (blue bars).
-Among the incidents with higher financial loss, there are more cases without fire alarm system.
+@fig-financial-loss-group shows the distribution of grouped economic loss and the presence of fire prevention system.
+In @fig-financial-loss-group-1, the graph shows that incidents with no fire alarm system (red bars) are less than incidents with fire alarm system (blue bars).
+Among the incidents with higher economic loss, there are more cases without a fire alarm system.
 In @fig-financial-loss-group-2, Most of the incidents happen with the smoke alarm present.
-Among the incidents with lower financial loss, more smoke alarms are present but not operated compared to higher financial loss cases.
-In @fig-financial-loss-group-3, we can tell that a large proportion of incidents does not have a sprinkler system installed.
+Among the incidents with lower economic loss, more smoke alarms are present but not operated compared to higher economic loss cases.
+In @fig-financial-loss-group-3, we can tell that a substantial proportion of incidents does not have a sprinkler system installed.
 Among those incidents with sprinkler system present, most of them are not operated.
@@
 
-The analysis of Toronto’s fire incident data reveals critical insights into both the responsiveness of the Toronto Fire Service (TFS) and the role of fire protection systems in mitigating financial losses.
+The analysis of Toronto’s fire incident data reveals critical insights into both the responsiveness of the Toronto Fire Service (TFS) and the role of fire protection systems in mitigating economic losses.
 
@@
 
-The Toronto Fire Service (TFS) generally has efficient response times, averaging about 4.9 minutes, as noted in the summary statistics [@tbl-clean-tfs-summary-1].
+The Toronto Fire Service (TFS) has efficient response times, averaging about 4.9 minutes, as noted in the summary statistics [@tbl-clean-tfs-summary-1].
 Over the period from 2013 to 2023, response times varied between 5 and 10 minutes, with some notable outliers, according to @fig-clean-tfs-diff-dollar-time-1.
 This indicates that while TFS usually responds quickly, there can be occasional delays due to outside factors.
-Importantly, these response times do not lead to significant financial losses, because the relationship between response time and financial impact is weak, as indicated by @fig-response-vs-money.
+Importantly, these response times do not lead to significant economic losses, because the relationship between response time and economical impact is weak, as indicated by @fig-response-vs-money.
 This implies that for the TFS, rapid response to fire incidents is not the most critical factor.
@@
 
-Financial losses from fire incidents vary greatly, with most cases resulting in low losses, as illustrated in @fig-clean-tfs-diff-dollar-time-2.
+Financial losses from fire incidents vary, with most cases resulting in low losses, as illustrated in @fig-clean-tfs-diff-dollar-time-2.
 However, certain outliers, especially from 2015 to 2024, have reported losses exceeding $2 million.
@@
 ## Impact of Fire Protection Equipment
-The presence and operation of fire protection equipment, particularly alarms and sprinklers, had a noticeable effect on financial losses.
-According to @fig-financial-loss-group, incidents where fire alarms were absent or failed to operate resulted in higher financial losses.
-This is consistent with previous researched carried out including the one conducted by @HariMurti2023 which further corroborates the studies that the absence of effective fire prevention systems makes the aftermath of fire destruction more damaging.
-SAdditionally, the data reveal that while smoke alarms were present in most cases, sprinklers were often absent or not operational.
+The presence and operation of fire protection equipment, particularly alarms and sprinklers, had a noticeable effect on economic losses.
+According to @fig-financial-loss-group, incidents where fire alarms were absent or failed to operate resulted in higher economic losses.
+This is consistent with previous research carried out including the one conducted by @HariMurti2023 which further corroborates the studies that the absence of effective fire prevention systems makes the aftermath of fire destruction more damaging.
+Additionally, the data reveals that while smoke alarms were present in most cases, sprinklers were often absent or not operational.
 This underlines the importance of incorporating sprinkler systems particularly in these areas where most fire outbreaks are recorded such as kitchens in accordance with statistics.
@@
 According to @fig-origin and @fig-source, kitchens and living rooms are the most common areas where fires start with most cases arising from stovetops and smokers’ articles being the most common.
-Such information makes it evident that fire outbreak should be prevented by targeting those measures in these severe areas.
-For instance fire incidents would be reduced considerable alongside its severity by making people understand the dangers of kitchen fires and the need to manage sprinkler and smoke alarm systems so that they do not malfunction.
+Such information makes it evident that fire outbreaks should be prevented by targeting those measures in these severe areas.
+For instance, fire incidents would be reduced considerably alongside its severity by making people understand the dangers of kitchen fires and the need to manage sprinkler and smoke alarm systems so that they do not malfunction.
 
@@
 
-This assessment is of great importance in the context of fire prevention and emergency response plans in the city of Toronto.
-First of all, the response times from TFS, apart from being fairly constant, also fluctuate at times, justifying the strategy that more improvements in the fire service will be in order in particular areas where there are delays.
-Furthermore, the data emphasize the great importance of fire protection systems in decreases in losses and the necessity of not only having fire protection systems in place but having them in working order as well.
+This assessment is important in the context of fire prevention and emergency response plans in Toronto.
+First, the response times from TFS, apart from being fairly constant, also fluctuate at times, justifying the strategy that more improvements in the fire service will be in order in particular areas where there are delays.
+Furthermore, the data emphasize the importance of fire protection systems in decreases in losses and the necessity of not only having fire protection systems in place but having them in working order as well.
 
-There are a few limitations on this research.
-First, there are other factors affecting TFS response time, such as local extreme weather or traffic.
+There are a few limitations to this research.
+First, there are other factors affecting TFS response time, such as extreme local weather or traffic.
 The response time data may not show the full picture of TFS's responsiveness.
 Second, the presence of fire protection equipment is oversimplified in the dataset.
-It is grouped to only three main categories.
-The state of fire alarm, smoke alarm and sprinkler system has to be analyzed more thoughtfully during the data collection process for a more useful analysis.
+It is grouped into only three main categories.
+The state of fire alarm, smoke alarm and sprinkler system have to be analyzed more thoughtfully during the data collection process for a more useful analysis.
 
-Overall, this analysis contributes to a deeper understanding of fire risks in urban environments and supports the development of targeted interventions aimed at minimizing fire-related damages.
+Overall, this analysis contributes to a deeper understanding of fire risks in urban environments and supports the development of targeted interventions aimed at minimizing fire-related damage.
 

### 2024-09-23 scripts/02-data_cleaning.r
fix data cleaning ignition source

@@ -179,7 +179,7 @@ cleaned_fire_data_simple <-
   cleaned_fire_data_simple |>
   mutate(
     ignition_source =
-      substr(ignition_source, start = 1, stop = 2)
+      substr(ignition_source, start = 1, stop = 3)
   )
 
 # Write the cleaned data to a new csv

### 2024-09-26 paper/paper.qmd
add data cleaning section suggested by colin in the peer review

@@
 
+The data cleaning process is recorded in @sec-datacleaning.
 The cleaned data are divided into two groups in accordance with the two most crucial factors in urban fire incidents: responsiveness of fire department and fire protection equipment, as indicated by @HariMurti2023 and @MasoodRafi2012.
@@
 
+## Data Cleaning {#sec-datacleaning}
+
+The data cleaning process starts with only keeping relavant data to this research.
+The first step keeps only the columns related to TFS response time, financial loss and the extenr or metadata of the fire incidents.
+Third, we identify and categorize all undetermined and invalid entries in the "Area of Origin" and "Ignition Source" columns that cannot be grouped.
+Forth, all columns related to time and money are converted to numeric value.
+The TFS response time is calculated by subtracting the fire incident report time from the TFS arrival time.
+Lastly, the categorical data are simplified to indicators such as "1", "12", "P" and "PO" to reduce the length in character of the dataset.
+
 ## Attribution Statement
